Introducing Distributed Dynamic Data-intensive (D3) Science: Understanding Applications and Infrastructure

### 2015-09-13 3dpas_intro.tex
spell check. partial.

The (grand) challenge in bioinformatics is the need to provide support
for the analysis of genome sequence data that is becoming available
due to the abundance and high-throughput capability of the
next-generation sequencing devices. The rate of growth of data from
the NGS machines humbles Moore's law; it has increased by more than
three orders of magnitude in barely a decade.

@@

The type of NGS research being conducted can distinguish the
distributed properties of the application. Researcher groups that
own sequencing machines need to deal with large dataset sizes that
need to be resequenced and reduced for further analysis. Other
researchers may analyze large catalogs of sequenced genome databases
are maintained, and perform pattern matching. Further, clinical may be
interested in in-depth analysis of an individual sequence for disease
diagnosis or visualization.

@@

There are several toolkits that have emerged for NGS providing
incrementally better algorithms for sequencing and analysis
\cite{SAMtools, BioJava, BioPerl, .NETBio}.  However, as a consequence
of the broad range of infrastructural requirements, there isn't
currently a common ``standard'' cyberinfrastructure used for NGS
analytics; researchers use what is easily available to them, or try to
use COTS infrastructure. Historically, leaders in the field have
developed infrastructure for the larger community to build-upon, e.g.,
QCDOC or MD-GRAPE. But given the broad range of requirements, a
hardware-only solution is unlikely for a broad range of problem
instances, and a combination of hardware and software approaches will
be employed. This reinforces the need for flexible programming systems
and associated runtime environments that support collective
(hardware-software) optimization.

Nowadays, NGS sequencers are even being shipped with servers or small
clusters for performing the alignment and resequencing as a pipeline
right after the reads have been generated~\cite{ilm}.
%\alnote{What kind of codes are used
%in these clusters? What are the characteristics of these codes?}
%But any additional analysis is not supported.
%\alnote{What would be an additional analysis}
%
Scientific workflow systems like Taverna~\cite{Taverna:2006}, BioKepler~\cite{biokepler:2011}, and
Galaxy~\cite{galaxy} (though Galaxy does not support distributed systems currently) have been successful in integrating genome
analysis tools for bioinformatics. Script based pipelines that combine
executables are also used often. Very few workflow systems have
explicit support for distributed data or dynamic data handling.
Therefore reusable approaches at extending existing tools to support
distributed and dynamic data are likely to be useful.
%\katznote{Shantenu/Yogesh/Andre, please read this paragraph and tell me if changes
%should be made to this section at this point}\alnote{ok with me}

@@

\item Using the existing object catalog (containing the known orbits of all known objects), find objects which are expected to appear in subtracted image, given the area of sky and time of day.
Cross-match expected objects with sources in subtracted image, resulting in the unmatched-source catalog, which contains sources that cannot be matched with a previously known object.  If these sources can be tracked over a few images, an orbit can be determined for them and they can be added to the object catalog.  Also, the orbit catalog can be updated with redetections of known objects; each rediscovery provides information that improves the known orbits.

@@

The application~\cite{modisAzure, modis:ipdps:2010} is used to handle the processes of downloading data from different
sites and integrating them. The integration consists of reprojecting the data from one map projection to another,
spatial and temporal resampling, and gap filling. The application itself is a loosely-coupled data
pipeline that coordinates these processes.

In addition, scientists can submit Matlab scripts for further custom processing on subsets of
the data that has been integrated. This includes some summarization processing and plotting
routines.  The only time constraints are on the debugging cycle for debugging the scientist-supplied
reduction codes. These are purely driven by the users' patience.

@@

{\bf MODIS Data Processing (\S\ref{modisKeith})}: Data are brought
into system from various FTP servers.  A pipeline of initial
standardized processing steps on data is done on clouds or HPC
resources.  Scientists can then submit executables and Matlab scripts
that do further custom processing on subsets of the data, which likely
include some summarization processing (building graphs).  The
application has been run on a cloud and in an HPC environment.

@@

{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express computations over 
collections of data which are executed in a loosely-coupled, data-parallel manner.
The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most
prominent among these, with those like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
graph analytics. MapReduce forces programmers to model computation in two stages: 
a \emph{map} phase in which a transformation is applied to every element of a collection independently of every other;  and a \emph{reduce} phase in which the transformed results 
are aggregated through a binary function that is associative and commutative. 
These properties serve to guarantee that the MapReduce computation can be efficiently 
parallelized, by performing the map phase concurrently and
then reducing without the reduction order affecting the final result.

@@

From \S\ref{sec:distTypes} and
\figurename~\ref{fig:figures_application}, we find that applications
with distributed data differ based on the amount or volume of data
distributed, the degree of distribution (scattered over one site or
many sites), and the dynamic aspects associated with distribution,
e.g., latency and relative scheduling flexibility (and thus tolerance
or constraints on data transfer/aggregation).  Applications processing
the largest volumes of data (O(PB)) typically run on specialized data
cyberinfrastructure (e.g., the Atlas LHC Grid). Applications such as
Fusion and NGS that run on smaller data sets mainly utilize HPC
environments. On the other end of the spectrum are highly dynamic
applications that process sensor data and make near-realtime
decisions based on streaming data, e.g., the power grid and sensor
network applications. These applications process much less data;
often, sensor data is also archived and aggregated for historical
analyses, which leads to higher data volumes. %\katznote{not sure about the `other' part of this...}

>>>>

The (grand) challenge in bio-informatics is the need to provide support
for the analysis of genome sequence data that is becoming available
due to the abundance and high-throughput capability of the
next-generation sequencing devices. The rate of growth of data from
the NGS machines humbles Moore's law; it has increased by more than
three orders of magnitude in barely a decade.

@@

The type of NGS research being conducted can distinguish the
distributed properties of the application. Researcher groups that
own sequencing machines need to deal with large dataset sizes that
need to be re-sequenced and reduced for further analysis. Other
researchers may analyze large catalogs of sequenced genome databases
are maintained, and perform pattern matching. Further, clinical may be
interested in in-depth analysis of an individual sequence for disease
diagnosis or visualization.

@@

There are several toolkits that have emerged for NGS providing
incrementally better algorithms for sequencing and analysis
\cite{SAMtools, BioJava, BioPerl, .NETBio}.  However, as a consequence
of the broad range of infrastructural requirements, there isn't
currently a common ``standard'' cyberinfrastructure used for NGS
analytics; researchers use what is easily available to them, or try to
use COTS infrastructure. Historically, leaders in the field have
developed infrastructure for the larger community to build-upon, e.g.,
QCDOC or MD-GRAPE. But given the broad range of requirements, a
hardware-only solution is unlikely for a broad range of problem
instances, and a combination of hardware and software approaches will
be employed. This reinforces the need for flexible programming systems
and associated run-time environments that support collective
(hardware-software) optimization.

Nowadays, NGS sequencers are even being shipped with servers or small
clusters for performing the alignment and re-sequencing as a pipeline
right after the reads have been generated~\cite{ilm}.
%\alnote{What kind of codes are used
%in these clusters? What are the characteristics of these codes?}
%But any additional analysis is not supported.
%\alnote{What would be an additional analysis}
%
Scientific workflow systems like Taverna~\cite{Taverna:2006}, BioKepler~\cite{biokepler:2011}, and
Galaxy~\cite{galaxy} (though Galaxy does not support distributed systems currently) have been successful in integrating genome
analysis tools for bio-informatics. Script based pipelines that combine
executable are also used often. Very few workflow systems have
explicit support for distributed data or dynamic data handling.
Therefore reusable approaches at extending existing tools to support
distributed and dynamic data are likely to be useful.
%\katznote{Shantenu/Yogesh/Andre, please read this paragraph and tell me if changes
%should be made to this section at this point}\alnote{ok with me}

@@

\item Using the existing object catalog (containing the known orbits of all known objects), find objects which are expected to appear in subtracted image, given the area of sky and time of day.
Cross-match expected objects with sources in subtracted image, resulting in the unmatched-source catalog, which contains sources that cannot be matched with a previously known object.  If these sources can be tracked over a few images, an orbit can be determined for them and they can be added to the object catalog.  Also, the orbit catalog can be updated with re-detections of known objects; each rediscovery provides information that improves the known orbits.

@@

The application~\cite{modisAzure, modis:ipdps:2010} is used to handle the processes of downloading data from different
sites and integrating them. The integration consists of re-projecting the data from one map projection to another,
spatial and temporal re-sampling, and gap filling. The application itself is a loosely-coupled data
pipeline that coordinates these processes.

In addition, scientists can submit Matlab scripts for further custom processing on subsets of
the data that has been integrated. This includes some summarizing processing and plotting
routines.  The only time constraints are on the debugging cycle for debugging the scientist-supplied
reduction codes. These are purely driven by the users' patience.

@@

{\bf MODIS Data Processing (\S\ref{modisKeith})}: Data are brought
into system from various FTP servers.  A pipeline of initial
standardized processing steps on data is done on clouds or HPC
resources.  Scientists can then submit executables and Matlab scripts
that do further custom processing on subsets of the data, which likely
include some summarizing processing (building graphs).  The
application has been run on a cloud and in an HPC environment.

@@

{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express
computations over collections of data which are executed in a loosely-coupled,
data-parallel manner.  The {\em MapReduce} paradigm~\cite{MapReduce} and its
variants~\cite{twister,hmr} are the most prominent among these, with those like
Pregel and Apache Giraph~\cite{pregel} supporting specific domains like graph
analytics. MapReduce forces programmers to model computation in two stages: a
\emph{map} phase in which a transformation is applied to every element of a
collection independently of every other; and a \emph{reduce} phase in which the
transformed results are aggregated through a binary function that is associative
and commutative.  These properties serve to guarantee that the MapReduce
computation can be efficiently parallelized, by performing the map phase
concurrently and then reducing without the reduction order affecting the final
result.

@@

From \S\ref{sec:distTypes} and
\figurename~\ref{fig:figures_application}, we find that applications
with distributed data differ based on the amount or volume of data
distributed, the degree of distribution (scattered over one site or
many sites), and the dynamic aspects associated with distribution,
e.g., latency and relative scheduling flexibility (and thus tolerance
or constraints on data transfer/aggregation).  Applications processing
the largest volumes of data (O(PB)) typically run on specialized data
cyberinfrastructure (e.g., the ATLAS LHC Grid). Applications such as
Fusion and NGS that run on smaller data sets mainly utilize HPC
environments. On the other end of the spectrum are highly dynamic
applications that process sensor data and make near-realtime
decisions based on streaming data, e.g., the power grid and sensor
network applications. These applications process much less data;
often, sensor data is also archived and aggregated for historical
analyses, which leads to higher data volumes. %\katznote{not sure about the `other' part of this...}
